Supplementary material to: Empirical Bayes estimation of the normal inverse Gaussian prior in simultaneous-equation models

### 2019-06-04 docs/supplement1.Rnw
first ENIG simulations

	\section{NIGIG prior}\label{sec:nigigprior}
	Equivalently, we may set 
  $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
  MD equation (\ref{md-eq:generalmodel}) as:
  \begin{align*}
    \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
    \(0, \frac{\lambda_{jd}}{\chi_{jd}}\eta_{jd}^2\sigma_d^2\), \\
    \eta_{jd}^2 & \overset{D}{=}\pi^{-1}\sqrt{\chi_{jd}}\exp(\psi_{jd}) 
    \eta_{jd}^{-3/2} \cdot K_{0} \(\sqrt{\psi_{jd}^2 + \chi_{jd} (\eta_{jd}^2 
    + \eta_{jd}^{-2} - 2)} \),
  \end{align*}
  where $\lambda_{jd}=\lambda_j^z\lambda_d^c$, 
  $\chi_{jd}=\lambda_j^z\lambda_d^c/(\theta_j^z \theta_d^c)$, and
  $\psi_{jd}=\lambda_j^z/\theta_j^z + \lambda_d^c/ \theta_d^c$. A third way
  of writing the model is:
  \begin{subequations}\label{eq:generalprior}
    \begin{align}
      \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
      \(0, \eta_{jd}^2 \sigma_d^2\), \\
      \eta_{jd}^2 | \xi_{d}^2 & \sim \mathcal{IG} 
      (\theta_j^z \xi_d^2, \lambda_j^z \xi_d^2 ), \\
      \xi_d^2 & \sim \mathcal{GIG} (1, \lambda_d^c/(\theta_d^c)^2, \lambda_d^c).
    \end{align}
  \end{subequations}
	From (\ref{eq:generalprior}), for a fixed error variance
	$\sigma_d^2$, we find the marginal prior variance and kurtosis:
	\begin{subequations}\label{eq:varkurtnigig}
	  \begin{align}
	    \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d^c \theta_j^z  
      \(r_d + \frac{2\theta_d^c}{\lambda_d^c}\), \\
	    \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \(\frac{3\theta_j^z}
	    {\lambda_j^z} + 3\) \frac{4 \theta_d^c (r_d + 2\theta_d^c/\lambda_d^c) +
	    1}{\lambda_d^c(r_d + 2\theta_d^c/\lambda_d^c)^2},
	  \end{align}
	\end{subequations}
	with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$.
	If 
  we fix the prior expectation $\E(\gamma_d^2)=\theta_d^c$ and let the prior 
  variance 
  $\V(\gamma_d^2)$ go to zero, we have $\V(\beta_{jd} | \sigma_d^2) \rightarrow
  \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:varkurtnigig}) we see
  that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
  and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
  $\lambda_j^z$. This makes the model MD equations (\ref{md-eq:generalmodel}) a
  flexible model.

@@

  \section{Efficient computation}
  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$. In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO. The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$. Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting. Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
  \begin{align*}
  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
  \end{align*}
  where $v_i$ are the singular values of $\X$. The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm. The rest of the calculations involve just $n$ scalar multiplications and/or additions.

>>>>

	\section{NIGIG prior}\label{sec:dnigprior}
	Equivalently, we may set 
  $\eta_{jd}^2=\tau_j^2 \gamma_d^2/(\theta_j^z \theta_d^c)$ and write 
  MD equation (\ref{md-eq:extendedmodel1}) as:
  \begin{align*}
    \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
    \(0, \frac{\lambda_{jd}}{\chi_{jd}}\eta_{jd}^2\sigma_d^2\), \\
    \eta_{jd}^2 & \overset{D}{=}\pi^{-1}\sqrt{\chi_{jd}}\exp(\psi_{jd}) 
    \eta_{jd}^{-3/2} \cdot K_{0} \(\sqrt{\psi_{jd}^2 + \chi_{jd} (\eta_{jd}^2 
    + \eta_{jd}^{-2} - 2)} \),
  \end{align*}
  where $\lambda_{jd}=\lambda_j^z\lambda_d^c$, 
  $\chi_{jd}=\lambda_j^z\lambda_d^c/(\theta_j^z \theta_d^c)$, and
  $\psi_{jd}=\lambda_j^z/\theta_j^z + \lambda_d^c/ \theta_d^c$. A third way
  of writing the model is:
  \begin{subequations}\label{eq:extendedmodel1.2}
    \begin{align}
      \beta_{jd} | \eta_{jd}^2, \sigma_d^2 & \sim \mathcal{N} 
      \(0, \eta_{jd}^2 \sigma_d^2\), \\
      \eta_{jd}^2 | \xi_{d}^2 & \sim \mathcal{IG} 
      (\theta_j^z \xi_d^2, \lambda_j^z \xi_d^2 ), \\
      \xi_d^2 & \sim \mathcal{GIG} (1, \lambda_d^c/(\theta_d^c)^2, \lambda_d^c).
    \end{align}
  \end{subequations}
	From (\ref{eq:extendedmodel1.2}), for a fixed error variance
	$\sigma_d^2$, we find the marginal prior variance and kurtosis:
	\begin{subequations}\label{eq:varkurtdgig}
	  \begin{align}
	    \V(\beta_{jd} | \sigma_d^2) & = \sigma_d^2 \theta_d^c \theta_j^z  
      \(r_d + \frac{2\theta_d^c}{\lambda_d^c}\), \\
	    \mathcal{K}(\beta_{jd} | \sigma_d^2) & = \(\frac{3\theta_j^z}
	    {\lambda_j^z} + 3\) \frac{4 \theta_d^c (r_d + 2\theta_d^c/\lambda_d^c) +
	    1}{\lambda_d^c(r_d + 2\theta_d^c/\lambda_d^c)^2},
	  \end{align}
	\end{subequations}
	with $r_d = K_0(\lambda_d^c/\theta_d^c)/K_1(\lambda_d^c/\theta_d^c)$.
	If 
  we fix the prior expectation $\E(\gamma_d^2)=\theta_d^c$ and let the prior 
  variance 
  $\V(\gamma_d^2)$ go to zero, we have $\V(\beta_{jd} | \sigma_d^2) \rightarrow
  \sigma_d^2 \theta_j^z \theta_d^c$ and $\mathcal{K}(\beta_{jd} | \sigma_d^2)
  \rightarrow 3(\theta_j^z/\lambda_j^z + 1)$. From (\ref{eq:varkurtdgig}) we see
  that we can keep the variance constant by fixing $\theta_j^z$, $\theta_d^c$,
  and $\lambda_d^c$, while we increasing the kurtosis by decrearing 
  $\lambda_j^z$. This makes the model MD equations (\ref{md-eq:extendedmodel1})
  a flexible model.

@@

  \section{Efficient computation}
  \subsection{NIG model}
  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$. In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO. The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$. Let $c_d^{(h)}=b_d^{(h)}$ in the conjugate setting and $c_d^{(h)}=b_d^{(h)}/a_d^{(h)}$ in the non-conjugate setting. Then, beforementioned quantities are easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
  \begin{align*}
  \trace (\bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \[\sum_{i=1}^n (v_i^2 + c_d^{(h)})^{-1} + \max (p - n, 0)\cdot (c_d^{(h)})^{-1}\], \\
  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = (a_d^{(h)})^{-1} \sum_{i=1}^n v_i^2(v_i^2 + c_d^{(h)})^{-1}, \\
  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)})^2, \\
  \log|\bm{\Sigma}^{(h+1)}| & \propto -p \log a_d^{(h)} - \sum_{i=1} \log (v_i^2 + c_d^{(h)}) - \max(p - n, 0) \cdot \log c_d^{(h)}, \\
  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + c_d^{(h)}).
  \end{align*}
  where $v_i$ are the singular values of $\X$. The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm. The rest of the calculations involve just $n$ scalar multiplications and/or additions.

  \subsection{DNIG and ENIG models}
  If we have the following forms:
  \begin{align*}
    \bm{\Sigma} = a^{-1} \(\tilde{\X} \tr \tilde{\X} + 
    \tilde{\mathbf{H}}\)^{-1}, \\
    \bm{\mu} = \(\tilde{\X} \tr \tilde{\X} + 
    \tilde{\mathbf{H}}\)^{-1} \tilde{\X} \tr \y,
  \end{align*}
  with $\tilde{\X} = \begin{bmatrix} \mathbf{1} & \X \end{bmatrix}$ and
  $$
  \tilde{\mathbf{H}} = 
  \begin{bmatrix} 
    0 & \mathbf{0} \\
    \mathbf{0} & \mathbf{H}
  \end{bmatrix},
  $$
  then 
  \begin{align*}
    \diag (\bSigma) = a^{-1} 
    \begin{bmatrix}
      q \\
      \mathbf{r}
    \end{bmatrix},
  \end{align*}
  with
  \begin{align*}
    \mathbf{D} & = (\I_n + \X \H^{-1} \X \tr), \\
    q & = \[ n - \mathbf{1} \X \H^{-1} \X \tr \mathbf{1} + 
    \mathbf{1} \X \H^{-1} \X \tr \mathbf{D}^{-1} 
    \X \H^{-1} \X \tr \mathbf{1}\]^{-1} \\
    \mathbf{r} & = \diag( \H^{-1}) - \left\{ \[\H^{-1} \X \tr 
    \mathbf{D}^{-1}\] \circ (\H^{-1} \X \tr) \right\} \mathbf{1} +
    q (\H^{-1} \X \tr \mathbf{1}) \circ (\H^{-1} \X \tr \mathbf{1}) \\
    & \,\,\,\,\, - 2q \[\H^{-1} \X \tr \mathbf{D}^{-1} 
    \X \H^{-1} \X \tr \mathbf{1}\] \circ (\H^{-1} \X \tr \mathbf{1}) \\
    & \,\,\,\,\, + q \[\H^{-1} \X \tr \mathbf{D}^{-1} 
    \X \H^{-1} \X \tr \mathbf{1}\] \circ \[\H^{-1} \X \tr 
    \mathbf{D}^{-1} \X \H^{-1} \X \tr \mathbf{1}\].
  \end{align*}
  Likewise we find 
  \begin{align*}
    \bm{\mu} =  
    \begin{bmatrix}
      q \sum_{i=1}^n y_i - q t + qs \\
      \mathbf{u}
    \end{bmatrix},
  \end{align*}
  with
  \begin{align*}
    s & = 1 \X \H^{-1} \X \tr \mathbf{D}^{-1} \X \tr \H^{-1}
    \X \tr \y \\
    t & = \mathbf{1} \X \H^{-1} \X \tr \y \\
    \mathbf{u} & = q \(t - \sum_{i=1}^n y_i - 2s \) \H^{-1} \X \tr \mathbf{1} + 
    + \H^{-1} \X \tr \y + q \( \sum_{i=1}^n y_i + s\) 
    \H^{-1} \X \tr \mathbf{D}^{-1} \X \H^{-1} \X \tr \mathbf{1} \\
    & \,\,\,\,\, - \H^{-1} \X \tr \mathbf{D}^{-1} \X \H^{-1} \X \tr \y.
  \end{align*}
  We have
  \begin{align*}
    \X \bmu & = q \( \sum_{i=1}^n y_i + s - t \) \mathbf{1} + 
    q \(t - \sum_{i=1}^n y_i - 2s \) \X \H^{-1} \X \tr \mathbf{1} +
    \X \H^{-1} \X \tr \y \\
    & \,\,\,\,\, + q \(s + \sum_{i=1}^n \) \X \H^{-1} \X \tr \mathbf{D}^{-1} 
    \X \H^{-1} \X \tr \mathbf{1} - \X \H^{-1} \X \tr \mathbf{D}^{-1} 
    \X \H^{-1} \X \tr \y.
  \end{align*}
  Furthermore
  \begin{align*}
    \bSigma \X \tr & = q \( \sum_{i=1}^n y_i + s - t \) \mathbf{1} + 
    q \(t - \sum_{i=1}^n y_i - 2s \) \X \H^{-1} \X \tr \mathbf{1} +
    \X \H^{-1} \X \tr \y \\
    & \,\,\,\,\, + q \(s + \sum_{i=1}^n \) \X \H^{-1} \X \tr \mathbf{D}^{-1} 
    \X \H^{-1} \X \tr \mathbf{1} - \X \H^{-1} \X \tr \mathbf{D}^{-1} 
    \X \H^{-1} \X \tr \y.
  \end{align*}
